---
title: "Converting Vectra data to spatialExperiment Objects"
hitheme: tomorrow
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
highlighter: highlight.js
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Overview 

This document provides info on how to convert Vectra data into a `spatialExperiment` object.

* Should have conversion functions for `.tiff` files and `.txt` tabular data
  * only `.txt` currently implemented

## Correspondence between ST and MI

Here I explain the correspondence between `spatialExperiment` slots for typical ST data, and how we expect it to behave for *multiplex imaging (MI) data.

* `assays`: contains expression counts
  * columns: cells
  * rows: genes
  * For *multiplex imaging data*:
    * In our case, columns are cells and rows are markers
    * Typically see 8-40 markers
    * For Vectra data, get `min/max/mean/total/std dev` expression, measured at `membrane/nucleus/total cell`, for a total of 15 variables for each marker
      * **Right now all of these are stored within one assay. The mean marker at the total cell level is most commonly used. Do we want to put this in one assay and store the rest in a second assay?**
    * In ST these are called `counts`, 
      * for MI I kept them as `counts` but want to rename `intensities`. Will this break everything?
* `rowData`: "feature metadata"
  * For us, this would contain information about markers... maybe marker type (phenotypic vs. functional)?
* `colData`: contains info about cells
  * For us this is phenotype, tissue category, cell shape, cell min/max axis, etc
* `spatialData`: Dataframe of spatial variables
  * `spatialCoords`: stores matrix of spatial coordinates
  * `in_tissue` argument: is pixel location tissue or slide? this can be determined from Vectra output.
* `imgData`: contains image data. Our images might be too big for this.


## Convert `.txt` to `spe`


I wrote a function called `readVectraTable.R` for converting Vectra `.txt` data to an `spe` object. You provide the function with a directory that has one or multiple `.txt` or `.txt.zip` files from a Vectra instrument and combines into one `spe` object.


### Lung cancer data

These data are 5 images from one subject. This is a subset of the data from [Johnson et. al.](https://www.semanticscholar.org/paper/Cancer-cell-specific-MHCII-expression-as-a-of-the-Johnson-Boland/310b4eb9d9fe2ec2254004e0780c4b722ba1fe5d) with 153 subjects and 3-5 images each.

```{r, echo = TRUE}
library(tidyverse)
library(SpatialExperiment)

source(here::here("source", "readVectraTable.R"))


spe = readVectraTable(sample_path = here::here("data", "tabular"),
                clean_names = TRUE)
```

### Phenoptr data

This data is available in the [`phenoptrExamples` package](https://akoyabio.github.io/phenoptrExamples/index.html) from Akoya Biosciences. Nine images taken from three samples of (also) lung cancer tissue.

```{r}
path <- system.file("extdata", "samples", package = "phenoptrExamples")

spe_phenoptr <- readVectraTable(sample_path = path)
```


## Downstream Analysis

Do some downstream analysis already possible with the `SpatialExperiment` package.

* multiplex imaging experiments will typically have more cells and fewer rows (genes) than a spatial transcriptomics experiment

```{r}
dim(spe)
```


```{r}
assayNames(spe)
```


* Prints out spatial data that is not the coordinates
  * What other variables might I want to put here?
  
  
```{r}
head(spatialData(spe)) 
```

* Print out spatial coordinates

```{r}
head(spatialCoords(spe))
```

### Visualization

Doesn't work- why not?

```{r, eval = FALSE}
library(ggspavis)
plotSpots(spe)


# example from vignette also doesn't work
spe_test <- SpatialExperiment(assay=mat, 
                         spatialData=cd, 
                         spatialCoordsNames=c("x", "y"))



plotSpots(spe_test)
```



### QC

```{r}
library(scater)

spe <- spe[, spatialData(spe)$in_tissue == 1]
dim(spe)
```



```{r}
library(scran)
set.seed(123)
qclus <- quickCluster(spe)
table(qclus)

spe <- computeSumFactors(spe, cluster = qclus)

summary(sizeFactors(spe))

spe <- logNormCounts(spe)
```

### My downstream analysis

* Convert to  cell populations in different areas (how to manipulate assays argument)
* Do log of mean expression values
* Would I need different methods? Which ones?

## Questions

* Right now we use `counts`, which is the same assay name as for ST objects. Would we lose functionality if we change this to `intensities` or something like that?
* What if I want to add in clinical data? Metadata data frame (hdcytodata)
* How do I want to organize `assays`? 
  * Min, mean, max, std dev, total all together or each as a separate assay?
  * cell, membrane, total all together or each as a separate assay?
  * the main assay object (counts) wouldn't be high dimensional, which is ok
* Not currently supporting storage of raw images- do you think this is fine? (What is the size?)
  * Downsampled images?


**Uploading Data to S3**

STExampleData- recently uploaded, has metadata scripts
* I already got access- what form should the data be in? An `RData` object?
  * Otherwise, could upload whole directory of data and do the processing as part of an annotation package
* devel version of Bioconductor bioc


## To Do

* Add rownames with subject-image-cell-id identifier
* Add image data?
* We may need to convert area characteristics from pixels to microns
  * See `phenoptr::read_cell_seg_data()` about this
* Build reference table for functional vs. surface vs immune marker types
* Make compatible with Cole's normalization. Add as new normalized assay?
* Add clinical data

### Methods

What methods will we want to make specifically for MI data?

* Normalization method based on Cole's package
  * `logNormCounts` (`scuttle` package) interacts with `SingleCellExperiment` object and adds a new assay to the object. We want something that functions similarly.
* Cell type summaries (does this already exist for `SingleCellExperiment`?)
* Nearest neighbor distances
* Distances to tissue area boundaries
* Moran's I, etc

### Images

I plan to write a function to convert images and segmentations to tabular format. Images might be too large for storage as part of a spatial experiment object.
