---
title: "Loading and Visualizing Vectra-Polaris Images"
hitheme: tomorrow
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
highlighter: highlight.js
---

# Overview

Below I load processed data from a Vectra instrument. We are looking at a tissue slice from a patient with lung cancer. Paper reference [here](https://www.biorxiv.org/content/biorxiv/early/2021/02/25/2021.02.24.432729.full.pdf).

The original analysis has 132 subjects with 3-5 images each. The dataset loaded has just one subject and three images.

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(phenoptr)

knitr::opts_chunk$set(echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

theme_set(theme_bw() + theme(legend.position = "bottom"))


```

# Load Filepaths

These images come from a lung cancer dataset with a published analysis [here](). There are roughly 132 subjects and each subjects have 3-5 images. 

The are collected on a Vectra instrument. Cell and tissue segmentation and cell phenotyping were performed using Inform software. For each image, there are multiple Tiff files:


* `*component_data.tif`: the multichannel file we are most interested in
* `*binary_sig*`: 3 channel file with binary segmentation maps
* `"subject_id".tif`: A composite image in RGB format. Makes pretty pictures.

```{r}
## load data
load(here("data", "vectra_table.Rdata"))

```

Unique images in the dataset:

```{r}
unique(vectra_table$`Sample Name`)
```

There are five images combined for this one subject.


# Analysis
I will add more simple analysis / visualization here
