---
title: "Loading and Visualizing Vectra-Polaris Images"
hitheme: tomorrow
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
highlighter: highlight.js
---

# Overview

Below I do some exploratory data analysis with multichannel tiff files from a Vectra instrument. We are looking at a tissue slice from a patient with lung cancer. Paper reference [here](https://www.biorxiv.org/content/biorxiv/early/2021/02/25/2021.02.24.432729.full.pdf).

I do some light image manipulation as well.

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
#library(rgdal)
#library(tiff)
library(phenoptr)

knitr::opts_chunk$set(echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

theme_set(theme_bw() + theme(legend.position = "bottom"))


```

# Load Filepaths

There are roughly 132 subjects and each subjects have 3-5 images. Only one image (there are multiple tiff files for the one image) from one subject is listed here.

The are collected on a Vectra instrument. Cell and tissue segmentation and cell phenotyping were performed using Inform software. For each image, there are multiple Tiff files:


* `*component_data.tif`: the multichannel file we are most interested in
* `*binary_sig*`: 3 channel file with binary segmentation maps
* `"subject_id".tif`: A composite image in RGB format. Makes pretty pictures.

```{r}
## define your specific file path
my_path = here("data")

# load csv filepaths
filenames = list.files(my_path)

component = paste0(my_path, "/", "#16 3-171-807_[62243,18509]_component_data.tif")
binseg = paste0(my_path, "/", "#16 3-171-807_[62243,18509]_binary_seg_maps.tif")
composite_tif = paste0(my_path, "/", "#16 3-171-807_[62243,18509].tif")
```


The subject and image ID are embedded in the file name. For example, for the file "#16 3-171-807_[62243,18509].tif, "#16 3-171-807" is the subject id, and "62243,18509" is the image ID. 


# Component files

The [`phenoptr` package](https://github.com/akoyabio/phenoptr) has functions for reading in data from Vectra.For the component files `phenoptr::read_components()` is a wrapper around `tiff::readTIFF()` which reads the individual component images from a `component_data.tif` file. It keeps the full-resolution component images, extracts the component names from the image descriptions, and returns a named list of image matrices.



Code below loads one multichannel tiff file. (I think) it returns a list of length 8 because there are 8 channels in this image. Names and dimensions of each channel are output below.

```{r}
image_stack = phenoptr::read_components(component)
map(image_stack, dim)

ls(image_stack)
```

Pull just the DAPI (nuclear) channel. This itself should be an image matrix.

```{r}
dapi_channel = image_stack$"DAPI (DAPI)"

dim(dapi_channel)
```


Component data may be displayed using `plot.raster or `EBImage::display`: 


```{r}
EBImage::display(t(dapi_channel))
```


# Segmentation files

The segmentation files are binary maps containing information about segmentation and either the tissue or cell level.
The `phenoptr::read_maps()` function is a wrapper around `readTIFF()` which reads segmentation files. It returns a named list of integer-valued matrices. The list names reflect the content of the individual images, e.g. Nucleus, Cytoplasm, etc.


```{r}

segmentations = phenoptr::read_maps(binseg)
names(segmentations)

```

In this case, we have segmentation maps of the nucleus, membrane, and tissue.


## Nucleus

```{r}

nucleus = segmentations[['Nucleus']]
dim(nucleus)
```

Below is a histogram of nucleus values. 

* 0 indicates background
* I guess we assume anything not 0 is part of a nucleus?
* I wonder if intensity indicates confidence that it is a real cell

```{r}
hist(nucleus)

table(nucleus[nucleus < 3])
```

Nuclei are plotted below:

```{r}

plot(as.raster(nucleus, max = max(nucleus)))
```

## Cytoplasm

Cytoplasm map is plotted below

```{r}
membrane = segmentations[['Membrane']]

plot(as.raster(membrane, max = max(membrane)))
```

## Tissue

Tissue map is plotted below

```{r}
tissue = segmentations[['Tissue']]

plot(as.raster(tissue, max = max(tissue)))
```

I believe 0 is stroma and 1 is tumor but it would be good to know this for sure. It would also be nice to overlay these images somehow.

```{r}
hist(tissue)
```


## image manipulation

### Extract healthy nuclei intensities

What I'd like to do next is apply both the nuclear and tissue masks to the dapi channel, so that areas that are not healthy (non-tumor) nucleus retain their original value and the rest is black

```{r, eval = FALSE}
dim(dapi_channel) 
dim(nucleus)
head(dapi_channel[,1:10])

```


* Modify nucleus segmentation map so that non-zero values are equal to 1. 
  * Later we should find out what the values actually mean in terms of certainty of the segmentation.
* Modify tissue map so that 0 is tumor and 1 is healthy tissue
  * We want to keep areas of healthy tissue
  * Also need to check what 0 and 1 mean biologically
  * Worth checking the CK channel to see where that tends to distribute

```{r}
nucleus[nucleus>0] = 1
#table(nucleus)

#table(tissue)
tissue = 1- tissue
#table(tissue)
```

* Now healthy tissue should take on a value of 1 in the tissue map and nuclei should take on a value of 1 in the nucleus map

```{r}
dapi_map = dapi_channel * nucleus * tissue
#dim(dapi_map)

plot(as.raster(tissue, max = max(tissue)))
plot(as.raster(dapi_map, max = max(dapi_map)))
```

What is the distribution of the nonzero `dapi_map` values?

```{r}
hist(dapi_map[dapi_map > 0])
```


# Resources

* Akoya provides some [info](https://akoyabio.github.io/phenoptr/articles/reading_image_files.html) on how to read their image files
